---
title: "Hubway Trip Data"
author: "Prasun Srivastava"
date: "27 March 2016"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---

The dataset for this analysis has been obtained from the [Hubway data Challenge.](http://hubwaydatachallenge.org/) The dataset contains the trip data for each bike ride from July 2011 to end of 2013. I will start with a few leading questions to guide my exploration:

* What is the distribution of the trip time? What is the average trip time?
* Do people use the bikes for shorter trip times or longer ones?
* When are most of the trips initiated? During what time of day, there are more commuters?
* Are there more male users or female ones? 
* What is the average age of the bike users? Does it differ among males and females?
* Which is the most popular station for source? Which station is most popular as destination? Which is the most popular source-destination pair?

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 12, 
                      fig.height = 8, 
                      fig.path = 'Figs/',
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r packages}
# Load all required packages
library(ggplot2)
library(dplyr)
library(lubridate)
library(ggmap)
```

Let us start by importing the dataset which is a csv file and viewing the structure of the data.

```{r}
trips <- read.csv('data\\hubway_trips.csv', stringsAsFactors = FALSE)
str(trips)
```
As can be seen from the structure of the dataframe, start and end stations are numeric values, start and end dates of the trips are strings which should be ideally date time variables. Next, we clean the data as per our requirement.

```{r}
trips$status     <- factor(trips$status)
trips$bike_nr    <- factor(trips$bike_nr)
trips$subsc_type <- factor(trips$subsc_type)
trips$zip_code   <- factor(sapply(trips$zip_code, substring, first = 2))
trips$gender     <- factor(trips$gender)
trips$start_date <- mdy_hms(trips$start_date, tz = 'EST')
trips$end_date   <- mdy_hms(trips$end_date, tz = 'EST')
stations         <- read.csv('data\\hubway_stations.csv', 
                             stringsAsFactors = FALSE)
station          <- stations$station
names(station)   <- as.character(stations$id)
trips$strt_statn_name <- station[as.character(trips$strt_statn)]
trips$end_statn_name  <- station[as.character(trips$end_statn)]
```

Now, let us start by summarising the data for each column.

```{r}
summary(trips)
```

The minimum trip duration is negative as seen above in the summary output! This is highly unlikely. This must be an error. Let us examine this record.

```{r}
trips[trips$duration == min(trips$duration), ]
```

The start date of the trip is greater than the end date of the trip which is not possible unless the person was driving a time machine!!! Let us remove all record from the trips data where the duration is negative.

```{r}
trips <- trips[trips$duration >= 0, ]
summary(trips)
```

Let us plot the distribution of trip duration.

```{r}
ggplot(data = trips, aes(x = duration)) + 
        geom_histogram(binwidth = 60, colour = 'black', fill = 'steelblue') +
        scale_x_continuous(limits = c(0, 4800)) +
        geom_vline(xintercept = median(trips$duration), colour = 'orange')
```

The trip duration is right skewed with most of the trips concentrated around the lower values. The median trip time (shown as orange vertical line)is 660s (11 minutes). 

Next, let us take a look at the trip times for registered users and casual users.

```{r}
ggplot(data = trips, aes(x = duration, fill = subsc_type)) + 
        geom_density(alpha = 0.3) +
        scale_x_continuous(limits = c(0, 4800)) +
        geom_vline(xintercept = 1800, 
                   colour = 'orange', 
                   linetype = 2, 
                   size = 1)
```

In the above plot, the vertical line depicts the free trip time of 30 minutes (1800s). Trips above 30 minutes incur overtime fees. More casual users tend to use the bike for more than 30 minutes than the registered users. Hence, the casual users must be incurring more overtime fees than the registered users. This might be due to the unawareness among casual users about the pricing scheme.

Next, let us take a look at the trip times for males and females. Since, we know that the gender information is captured only for the registered members, let us split our trips data based on membership.

```{r}
ggplot(data = subset(trips, subsc_type == 'Registered'), aes(x = duration)) + 
        geom_histogram(binwidth = 60, colour = 'black', fill = 'steelblue') +
        scale_x_continuous(limits = c(0, 4800)) +
        geom_vline(xintercept = median(trips$duration), colour = 'orange') +
        facet_wrap(~gender)
by(subset(trips, subsc_type == 'Registered')$duration, 
   subset(trips, subsc_type == 'Registered')$gender, median)
```

The distribution of trip times for both males and females is similar except the median trip time for females is slightly higher than the males. This might be due to the variation in speed of riding bikes among males and females. We will try to explore this later on. Also, from the count of the trips, it seems that there are more males using the bike service than the females.

Next, let us look at the distribution of trips over different days of week.

```{r}
trips$day_of_week <- wday(trips$start_date, label = TRUE)
ggplot(data = trips, aes(x = day_of_week)) + geom_bar(fill = 'steelblue')
```

From the above plot, we can see that the number of trips are similar for weekdays and low on weekends. This might be due to the fact that the people use bikes for commute to offices and there is no commute to work on weekends. However, if that were the case, the number of trips on weekends should have been far lower than the weekdays; which is not the case. 

The above plot has got me interested. People who are registered users might be the ones who use the bike service to commute on a daily basis. Does the subscription type affect the trips distribution by day?

```{r}
ggplot(data = trips, aes(x = day_of_week)) + 
        geom_bar(fill = 'steelblue') +
        facet_wrap(~subsc_type)
```

The registered users use the bike service more on weekdays and the casual users more on weekends. One possible explanation to this might be that the registered users use the bike service to commute to work on a daily basis whereas the casual users are the ones who roam the city for fun or maybe they are tourists.

Next, I want to look if male and females have different patterns for bike use on different days of week.

```{r}
ggplot(data = subset(trips, subsc_type == 'Registered'), 
       aes(x = day_of_week)) +
        geom_bar(fill = 'steelblue') +
        facet_wrap(~gender)
```

There is no apparent pattern. Both males and females have similar bike use pattern on all days of the week. Next, let us explore the trips by time of day.

```{r}
trips$hour_of_day <- hour(trips$start_date)
ggplot(data = trips, aes(x = hour_of_day)) + 
        geom_histogram(fill = 'steelblue', colour = 'black', binwidth = 1)
```

The bike usage peaks two times during the day; once between 8 and 9 AM in the morning and again between 5 and 6 PM in the evening. This again suggests that the bike users might be commuting to and from office. Next, let us look at the distribution by hour for each day of the week.

```{r}
ggplot(data = trips, aes(x = hour_of_day)) + 
        geom_histogram(fill = 'steelblue', colour = 'black', binwidth = 1) +
        facet_wrap(~day_of_week)
```

Our assumption about bike ride to and from office might be correct as the same pattern holds for weekdays but not on weekends. On weekends, the data is mostly normally distributed peaking around 3 PM in the afternoon. Moreover, the number of trips taken on Saturday is more than on Sunday. It seems people want to spend time at home on Sundays. Also, if we look closely, we find that the bike trips rise on Saturday and Sunday between 12 AM and 3 AM as compared to other days. This might be due to people going out on Fridays and Saturdays at night. 

Next, let us look at the same distribution for casual and registered users.

```{r}
ggplot(data = trips, aes(x = hour_of_day)) + 
        geom_histogram(fill = 'steelblue', colour = 'black', binwidth = 1) +
        facet_grid(subsc_type~day_of_week)
```

Again, it is evident that the casual users utilize bike services more during the weekends than weekdays. However, I want to summarize all the findings in a single chart. As weekdays have overall similar pattern, I will split the data as per weekday and weekend.

```{r}
calc_wkday_wkend <- function(day){
        if (day %in% c('Mon', 'Tues', 'Wed', 'Thurs', 'Fri'))
           return('Weekday')
        else if (day %in% c('Sat', 'Sun'))
           return('Weekend')
}
trips$day <- unname(sapply(as.character(trips$day_of_week), calc_wkday_wkend))
ggplot(data = trips, aes(x = hour_of_day, colour = subsc_type)) + 
        geom_density(adjust = 2, aes(y = ..count..)) + facet_wrap(~day)
```

How does the bike usage differ by month?

```{r}
trips$month <- month(trips$start_date, label = TRUE)
ggplot(data = trips, aes(x = month)) + 
        geom_bar(fill = 'steelblue', colour = 'black')
```

The bike usage peaks in August and September and there is no usage during the months of December, January and February. This makes sense as December, January and February are the coldest months of the year.

Next, I want to understand the age distribution of the people riding the bikes. Since, we have birth date information only for the registered users, we will plot the data for them only.

```{r}
trips$age <- year(trips$start_date) - trips$birth_date
ggplot(data = subset(trips, subsc_type == 'Registered'),
       aes(x = age)) + 
        geom_histogram(fill = 'steelblue', colour = 'black', binwidth = 2) +
        scale_x_continuous(breaks = seq(10, 70, 2), limits = c(10, 70))
```

The age distribution is also right skewed with most people under the ages of 35. Let us take a look at the distribution of the ages by gender.

```{r}
ggplot(data = subset(trips, subsc_type == 'Registered'),
       aes(x = age)) + 
        geom_density(aes(fill = gender, y = ..count..), alpha = 0.5)
by(trips$age, trips$gender, summary)
```

The age distribution for both males and females is right skewed with median ages being 31 and 33 year for females and males respectively.

Let us try to estimate the speed of the bike. We have duration of the trip and if we are able to calculate the distance between the two stations, we will be able to estimate the speed. In order to do this, I compiled the distance between two stations using the Google maps distance API in a csv file (`origin_destination.csv`). Since, there is an upper limit on 2500 queries per day for the API, the data was compiled over a period of time for around 14000 unique trips. A sample script used to compile the distances is in the project folder (`get_distance.R`). Also, for this analysis, only those trips will be considered where source and destination were different and at least one of the station info is present.

```{r}
distance <- read.csv('data\\origin_destination.csv')
station_lat <- stations$lat
names(station_lat) <- as.character(stations$id)
station_lon <- stations$lng
names(station_lon) <- as.character(stations$id)
trips$strt_lat <- station_lat[as.character(trips$strt_statn)]
trips$strt_lon <- station_lon[as.character(trips$strt_statn)]
trips$end_lat <- station_lat[as.character(trips$end_statn)]
trips$end_lon <- station_lon[as.character(trips$end_statn)]

# perform inner join to consider distances only between source-destination pairs
# present in the origin_destination file.

trips_distance <- trips %>% 
                   inner_join(distance, 
                     by = c('strt_lat', 'strt_lon', 'end_lat', 'end_lon')) %>%
                   select(seq_id,
                          hubway_id, 
                          status,
                          duration, 
                          start_date, 
                          strt_statn,
                          end_date, 
                          end_statn,
                          bike_nr, 
                          subsc_type, 
                          zip_code, 
                          birth_date,
                          gender, 
                          strt_statn_name, 
                          end_statn_name, 
                          day_of_week,
                          hour_of_day,
                          day, 
                          month, 
                          age, 
                          strt_lat, 
                          strt_lon, 
                          end_lat,
                          end_lon, km)

# remove trips with duration 0s
trips_distance <- subset(trips_distance, duration != 0)

# divide duration by 3600 to convert in hour. Speed calculated in km/hr
trips_distance$speed <- trips_distance$km / (trips_distance$duration / 3600)

# plot the speed distribution
ggplot(data = trips_distance, aes(x = speed)) +
        geom_histogram(fill = 'steelblue', colour = 'black', binwidth = 5) +
        scale_x_continuous(limits = c(0, 80))
```

Most of the trips are estimated to be between speeds of 10 - 15 Km/hr. Although, there are many trips that have an estimated speed grater than 60 Km/hr which seems very unlikely to me. Let us take a look at the speed distribution for males and females by age.

```{r}
ggplot(data = subset(trips_distance, subsc_type == 'Registered'), 
       aes(x = age, y = speed)) + 
        geom_smooth(aes(colour = gender), se = FALSE) 
```

Older people tend to cycle slower than the younger people and males tend to cycle faster on average than females. Let us check if the subscription type affects the speed.

```{r}
ggplot(data = trips_distance, 
       aes(x = speed)) + 
        geom_density(aes(y = ..count.., colour = subsc_type)) +
        scale_x_continuous(limits = c(0, 40))
```

The registered users tend to have higher speeds than the casual bikers. This might be because casual bikers might be using to roam around leisurely. Next, let us check the distribution of the distance.

```{r}
ggplot(data = trips_distance, aes(x = km)) +
        geom_histogram(fill = 'steelblue', colour = 'black', binwidth = 1)
```

Next, let us look at the distribution of distance by gender and examine the distance summary by gender.

```{r}
ggplot(data = subset(trips_distance, subsc_type == 'Registered'), 
       aes(x = km)) +
        geom_histogram(fill = 'steelblue', colour = 'black', binwidth = 1)+
        facet_wrap(~gender)
by(subset(trips_distance, subsc_type == 'Registered')$km, 
   subset(trips_distance, subsc_type == 'Registered')$gender, summary)
```

Earlier, we had seen that the trip duration for females is slightly higher than the males. However, this is due to the reason that the females have a higher median distance traveled.

Let us move forward to answer the question about most popular source station.

```{r}
# group trips by start station name and count number of trips for each station
tripsby_strt_statn <- trips %>%
                         group_by(strt_statn_name) %>%
                         summarise(number_trips = n()) %>%
                         select(strt_statn_name, number_trips) %>%
                         arrange(desc(number_trips)) %>%
                         ungroup()
tripsby_strt_statn$strt_statn_name <- factor(tripsby_strt_statn$strt_statn_name, 
                levels = tripsby_strt_statn$strt_statn_name, ordered = TRUE)
ggplot(data = tripsby_strt_statn[1:5, ], 
       aes(x = strt_statn_name, y = number_trips)) +
        geom_bar(fill = 'steelblue', stat = 'identity') +
        theme(axis.text.x = element_text(size = 8))
```

In the above plot, we have plotted the top 5 source stations. The Seaport Boulevard is the top source station. However, now I have a question; does the top source stations change during different times of day? In order to answer this question, we would divide whole day in 4-hour buckets. Let us define these buckets as follows:

* 12 AM - 4 AM: Night
* 4 AM - 8 AM: Early morning
* 8 AM - 12 PM: Morning
* 12 PM - 4 PM: Afternoon
* 4 PM - 8 PM: Evening
* 8 PM - 12 AM: Late Evening

```{r}
create_bucket <- function(hr){
        if (hr %in% c(0:3)) return('Night')
        else if (hr %in% c(4:7)) return('Early Morning')
        else if (hr %in% c(8:11)) return('Morning')
        else if (hr %in% c(12:15)) return('Afternoon')
        else if (hr %in% c(16:19)) return('Evening')
        else if (hr %in% c(20:23)) return('Late Evening')
}
trips$start_bucket <- sapply(trips$hour_of_day, create_bucket)
strt_stn_period_day <- trips %>%
                         group_by(strt_statn_name, start_bucket) %>%
                         summarise(number_trips = n()) %>%
                         select(strt_statn_name, 
                                start_bucket, 
                                number_trips) %>%
                         ungroup() %>%
                         group_by(start_bucket) %>%
                         top_n(n = 5, wt = number_trips) 
morning <- strt_stn_period_day[strt_stn_period_day$start_bucket == 'Morning', ]
evening <- strt_stn_period_day[strt_stn_period_day$start_bucket == 'Evening', ]
ggplot(data = morning, 
       aes(x = factor(strt_statn_name, 
                      levels = strt_statn_name[order(number_trips, decreasing =  TRUE)], ordered = TRUE), y = number_trips)) + 
        geom_bar(stat = 'identity', fill = 'steelblue') + 
        labs(x = 'Station Name', y = 'Number of Trips', title = 'Morning')
ggplot(data = evening, 
       aes(x = factor(strt_statn_name, 
                      levels = strt_statn_name[order(number_trips, decreasing =  TRUE)], ordered = TRUE), y = number_trips)) + 
        geom_bar(stat = 'identity', fill = 'steelblue') + 
        labs(x = 'Station Name', y = 'Number of Trips', title = 'Evening')
```

As can be seen from the above plot, the most popular source stations change between morning and evening except for the first most popular station (South Station 700 Atlantic Avenue). 

This has got me thinking what are the most popular source-destination pairs by different times of day.

```{r}
srcdest_period_day <- trips %>%
                       group_by(strt_statn_name, 
                                strt_statn, 
                                end_statn_name, 
                                end_statn, 
                                start_bucket) %>%
                       summarise(number_trips = n()) %>%
                       select(strt_statn_name, 
                              strt_statn, 
                              end_statn_name, 
                              end_statn, 
                              start_bucket, 
                              number_trips) %>%
                       ungroup() %>%
                       group_by(start_bucket) %>%
                       top_n(n = 5, wt = number_trips) %>%
                       arrange(start_bucket, desc(number_trips))

# display data as table
as.data.frame(select(srcdest_period_day, 
                     start_station = strt_statn_name,
                     end_station = end_statn_name,
                     day_period = start_bucket,
                     number_trips))
```

Although the above table output is okay to understand the top 5 source-destination pairs by time of day, it would be nice if we could plot the above data on top of a map. In order to do this, I compiled the bicycle routes for each source-destination pair from Google maps route API into a single `routes_info.csv` file. The script used to generate the file is included in the project folder. The name of the file is `get_route.R`. Let us plot the route of these source-destination pairs on the map.

```{r}
srcdest_period_day <- trips %>%
                        group_by(strt_statn_name, 
                                 strt_statn, 
                                 end_statn_name, 
                                 end_statn, 
                                 start_bucket) %>%
                        summarise(number_trips = n()) %>%        
                        select(strt_statn_name, 
                               strt_statn, 
                               end_statn_name, 
                               end_statn, 
                               start_bucket, 
                               number_trips) %>%
                        ungroup() %>% 
                        group_by(start_bucket) %>%
                        top_n(n = 500, wt = number_trips) %>%
                        arrange(start_bucket, number_trips)
strt_stn_index <- as.character(srcdest_period_day$strt_statn)
end_stn_index <- as.character(srcdest_period_day$end_statn)
srcdest_period_day$strt_lat <- station_lat[strt_stn_index]
srcdest_period_day$strt_lon <- station_lon[strt_stn_index]
srcdest_period_day$end_lat <- station_lat[end_stn_index]
srcdest_period_day$end_lon <- station_lon[end_stn_index]
route_df <- read.csv('data\\routes_info.csv', stringsAsFactors = FALSE)
morning <- srcdest_period_day[srcdest_period_day$start_bucket == 'Morning',]
evening <- srcdest_period_day[srcdest_period_day$start_bucket == 'Evening',]
afternoon <- srcdest_period_day[srcdest_period_day$start_bucket == 'Afternoon',]
early_mrng <- srcdest_period_day[srcdest_period_day$start_bucket == 'Early Morning',]
late_eve <- srcdest_period_day[srcdest_period_day$start_bucket == 'Late Evening',]
night <- srcdest_period_day[srcdest_period_day$start_bucket == 'Night',]
morning <- morning %>%
             inner_join(route_df, by = c('strt_lat',
                                         'strt_lon',
                                         'end_lat',
                                         'end_lon')) %>%
             arrange(number_trips, pathID, leg)

evening <- evening %>%
             inner_join(route_df, by = c('strt_lat',
                                         'strt_lon',
                                         'end_lat',
                                         'end_lon')) %>%
             arrange(number_trips, pathID, leg)

afternoon <- afternoon %>%
             inner_join(route_df, by = c('strt_lat',
                                         'strt_lon',
                                         'end_lat',
                                         'end_lon')) %>%
             arrange(number_trips, pathID, leg)

early_mrng <- early_mrng %>%
               inner_join(route_df, by = c('strt_lat', 
                                           'strt_lon', 
                                           'end_lat',
                                           'end_lon')) %>% 
               arrange(number_trips, pathID, leg)
                                            
late_eve <- late_eve %>%
                inner_join(route_df, by = c('strt_lat', 
                                            'strt_lon', 
                                            'end_lat', 
                                            'end_lon')) %>% 
                arrange(number_trips, pathID, leg)
                                             
night <- night %>%
           inner_join(route_df, by = c('strt_lat', 
                                       'strt_lon', 
                                       'end_lat',
                                       'end_lon')) %>%
           arrange(number_trips, pathID, leg)
                                           
boston <- qmap('boston', zoom = 14, color = 'bw')

# create_map takes a data frame for time of day and plots the routes on a map
create_map <- function(data_frame){
        boston + geom_path(data = data_frame,
                   aes(x = startLon,
                       y = startLat,
                       group = factor(pathID),
                       alpha = number_trips),
                   size = 1,
                   colour = '#6495ED') +
                coord_map()
}

# plot early morning bike ride routes
create_map(early_mrng)+ labs(title = 'Early Morning')

# plot morning bike ride routes
create_map(morning)+ labs(title = 'Morning')

# plot afternoon bike ride routes
create_map(afternoon)+ labs(title = 'Afternoon')

# plot evening bike ride routes
create_map(evening)+ labs(title = 'Evening')

# plot late evening bike ride routes
create_map(late_eve)+ labs(title = 'Late Evening')

# plot night bike ride routes
create_map(night)+ labs(title = 'Night')

```

# Final Plots and Summary

In this section, I will discuss the three main takeaways from this data exploration.

### First Plot

The first plot that I have chosen as the key takeaway is the plot of trip times.

```{r}
ggplot(data = trips, aes(x = duration, fill = subsc_type)) + 
        geom_density(aes(y = ..count..), 
                     alpha = 0.3) +
        scale_x_continuous(limits = c(0, 4800)) +
        geom_vline(xintercept = 1800, 
                   colour = 'orange', 
                   linetype = 2, 
                   size = 1) +
        labs(x = 'Trip Duration (in seconds)', 
             y = 'Number of Trips', 
             title = 'Trip Duration by Subscription Type')
```

The casual users tend to use the bike beyond the free usage limit of 30 minutes. This incurs overtime fee for these users. This might be due to lack of awareness among the casual users about the pricing of bike use. Hubway should raise awareness among the casual users about their pricing scheme.

### Second Plot

The second plot shows the distribution of trips by time of day.

```{r}
ggplot(data = trips, aes(x = hour_of_day)) + 
        geom_histogram(fill = 'steelblue', colour = 'black', binwidth = 1) +
        facet_grid(subsc_type~day_of_week) +
        labs(x = 'Hour of Day',
             y = 'Number of Trips',
             title = 'Trip variation during Day by Subscription Type')
```

The casual users use the bike service mostly on weekend whereas the registered users use the service mostly during weekdays. During the weekdays, the bike use among the registered users shows two clear peaks in usage: one during 8 AM to 9 AM and the other between 5 PM to 6 PM. This is suggestive of bike use for commute to and from office. There is a rise in bike use on Saturday and Sunday by both casual and registered users between 12 AM to 3 AM as compared to other days. This might be due to people going out on Friday and Saturday night. 

### Third Plot

The third plot shows the bike routes by number of trips 

```{r}
# plot early morning bike ride routes
create_map(early_mrng) + labs(title = '4 AM - 8 AM')

# plot morning bike ride routes
create_map(morning) + labs(title = '8 AM - 12 PM')

# plot afternoon bike ride routes
create_map(afternoon) + labs(title = '12 PM - 4 PM')

# plot evening bike ride routes
create_map(evening) + labs(title = '4 PM - 8 PM')

# plot late evening bike ride routes
create_map(late_eve) + labs(title = '8 PM - 12 AM')

# plot night bike ride routes
create_map(night) + labs(title = '12 AM - 4 AM')
```

The area around Boylston Street, Arlington Street and Back Bay is active during all times. People tend to visit the harborwalk area around the New England Aquarium during early morning and morning hours. There are a lot of trips between South Station and Atlantic Avenue during all times except during night. Also, in the evening, there is activity between South Station and TD Garden. 

# Reflection

I started off with a set of questions to guide my exploration and was able to dive into those questions during the analysis. As a person who has never been to Boston, the bike use data has provided with great insights into the culture of the city. There are patterns which suggest the use of bike for commute during weekdays. People seem to spend time during early hours of the day around the waterfront areas. Although these conclusions are not certain and might not be representative of the culture of the city as a whole but it still has provided me with some insights to the city.

Most of the analysis was done pretty quickly. However, the compilation of distances and routes from the Google maps API took a lot of time. In order to plot the routes on the map, I first started off with plotting greater circle arcs on a blank outline of Boston. However, it was soon clear that it was not a viable solution and did not provide any insights into the bike use pattern. Then, I started by plotting a few routes on the map of Boston and it was clear that it was the  way to go about finding patterns in bike routes during different times of day. Having no experience in plotting maps, it took me a lot of time to understand the basics of spatial visualization and then move towards the end goal that I had in vision.